Sampling requirements to detect ecosystem change

### 2020-10-05 R/supp_material.Rmd
edits to address reviewer comments

source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)
require(stickylabeller)

@@

require(ggplot2)
require(viridis)
ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
  facet_wrap(~shift_size_vec) + 
  theme_classic(base_size = 14, base_family = "")+
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(fill='Power',subtitle = 'Shift size') +
  xlab("Lag-1 autocorrelation (phi)") +
  ylab("Standard deviation (sigma)")
```

@@

>>>>

source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)
require(stickylabeller)
require(ggplot2)
require(viridis)

@@

ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
  facet_wrap(~shift_size_vec) + 
  theme_classic(base_size = 14, base_family = "")+
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(fill='Power',subtitle = 'Shift size') +
  xlab("Lag-1 autocorrelation (phi)") +
  ylab("Standard deviation (sigma)")
```

@@

# Case study: Testing other types of models

Throughout the main text and supplementary material, we focus on detecting changepoints. However, other types of models and criteria may be more appropriate depending on the question. This is especially true for situations where changepoints are not expected apriori. To address this, we used our random and regular sampling approaches with two additional models: linear models and generalized additive models. For simple linear regression, we measured the percent difference in the estimated slope between the subsampled data and the full time series (Fig. \ref{fig:lineartrends}). This is similar to past work [@White2019e; @White2020] on detecting population trends.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling. Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples. \\label{fig:lineartrends}'}

vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
vr_count <- data.frame(Year=vr[,1], Biomass=rowSums(vr[,-1]))
cpvar_count <- sample_regular(vr_count[,2], nrow(vr_count), input_vector = T, messages = F)

#plot(vr_count)
#abline(lm(vr_count$Biomass~vr_count$Year))
true_slope <- as.numeric(coef(lm(vr_count$Biomass~vr_count$Year))[2])

regular_linear <- function(x,y,num_samples){
  true_slope <- as.numeric(coef(lm(y~x))[2])

  subset_index <- c(seq(from = 1, to = length(x),length.out =num_samples),length(x))
  subset_x <- x[subset_index]
  subset_y <- y[subset_index]

  subset_slope <- as.numeric(coef(lm(subset_y~subset_x))[2])

 return(abs(true_slope - subset_slope)/true_slope)  
}

spacing_vec <- seq(3,70,by=3)
distance <- vector(mode='numeric',length = length(spacing_vec))
for (i in 1:length(spacing_vec)){
  distance[i] <- regular_linear(x = vr_count$Year,y=vr_count$Biomass,num_samples =spacing_vec[i])

}
par(mfrow=c(1,2),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))
plot(spacing_vec,distance,las=1,ylim=c(0,1.2),col='darkgrey',pch=16)
abline(h=0,col='darkgrey',lwd=1.5)
mtext('a. Random sampling - linear model',side = 3,line = 0,adj = 0)

# Random sampling

random_linear <- function(x,y,num_samples){
  true_slope <- as.numeric(coef(lm(y~x))[2])

  subset_index <- sort(c(1,sample(2:(length(x)-1),size = num_samples),length(x)))
  subset_x <- x[subset_index]
  subset_y <- y[subset_index]

  subset_slope <- as.numeric(coef(lm(subset_y~subset_x))[2])

  return(abs(true_slope - subset_slope)/true_slope)  
}

#spacing_vec <- 3:60
distance <- vector(mode='numeric',length = length(spacing_vec))
for (i in 1:length(spacing_vec)){
  distance[i] <- random_linear(x = vr_count$Year,y=vr_count$Biomass,num_samples =spacing_vec[i])

}
plot(spacing_vec,distance,las=1,ylim=c(0,1.2),yaxt='n',col='darkgrey',pch=16)
abline(h=0,col='darkgrey',lwd=1.5)
mtext('b. Regular sampling - linear model',side = 3,line = 0,adj = 0)
mtext('Percent difference',side = 2,line = 2.5,outer = T,font = 2)
mtext('Total number of samples analyzed',side = 1,line = 2.5,outer = T,font = 2)

Similarly, for a generalized additive model, we examined the shape of the estimated curve for subsets of data compared to the full time series (Fig. \ref{fig:gamtrends}). In line with our other results, the regular sampling performed better than random sampling upon visual inspection. Both sampling approaches also visually approach the full data set with fewer samples compared to the linear models.

```{r echo=FALSE, eval=T,message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Percent difference for the estimated slope from linear regression for the abundance time series using (a) random and (b) regular sampling. Iterative sampling could be used for linear trends, but would require additional decisions on how to add samples. \\label{fig:gamtrends}'}

vr <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt"))
vr_count <- data.frame(Year=vr[,1], Biomass=rowSums(vr[,-1]))
cpvar_count <- sample_regular(vr_count[,2], nrow(vr_count), input_vector = T, messages = F)

smoothing_parameter=5
require(mgcv)

par(mfrow=c(2,4),mar=c(0.5,0.1,1.5,0.1),oma=c(4,4,0.5,0.5))

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 10),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam1_rand,ylim=c(-50000,50000),xaxt='n')
mtext('a. Random (n=10)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 20),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam2_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('b. Random (n=20)',side = 3,line = 0,adj = 0)

subset_index <- sort(c(1,sample(2:(length(vr_count$Year)-1),size = 30),length(vr_count$Year)))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam3_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('c. Random (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot(gam4_rand,ylim=c(-50000,50000),xaxt='n',yaxt='n')
mtext('d. Random (n=74)',side = 3,line = 0,adj = 0)

# Regular GAMs

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =10),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam1_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam1_rand,ylim=c(-50000,50000))
mtext('e. Regular (n=10)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =20),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam2_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam2_rand,ylim=c(-50000,50000),yaxt='n')
mtext('f. Regular (n=20)',side = 3,line = 0,adj = 0)

subset_index <- c(seq(from = 1, to = length(vr_count$Year),length.out =30),length(vr_count$Year))
vr_count_subset <- vr_count[subset_index,]
gam3_rand <- gam(vr_count_subset$Biomass ~ s(vr_count_subset$Year,k=smoothing_parameter),family=gaussian,)
plot(gam3_rand,ylim=c(-50000,50000),yaxt='n')
mtext('g. Regular (n=30)',side = 3,line = 0,adj = 0)

gam4_rand <- gam(vr_count$Biomass ~ s(vr_count$Year,k=smoothing_parameter),family=gaussian)
plot(gam4_rand,ylim=c(-50000,50000),yaxt='n')
mtext('h. Regular (n=40)',side = 3,line = 0,adj = 0)

mtext('Standardized abundance',side = 2,line = 2.5,outer = T,font = 2)
mtext('Time (year)',side = 1,line = 2.5,outer = T,font = 2)
